A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-08 tex/codatadp.tex
Clarify a few points.

@@

A \emph{derivation tree} is a tree whose nodes are sets of equalities, where non-root nodes are obtained by 
a non-redundant application of a derivation rule to its parent node.
A derivation tree is \emph{closed} if all of its leaf nodes are $\bot$.
A node is \emph{saturated} if no non-redundant application of a rule can be applied to it.

\begin{lemma}[Termination]
All derivation trees are finite.
\end{lemma}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S$ be the set of terms occurring as the argument of a selector in $\tEc$.
Let $T$ be the set of terms in $\tEc$ of finite datatype type.
For each term $t \in T$,
let $S^0_t$ be the set $\{ t \}$,
for $i \geq 0$ let $S^{i+1}_t =$ $S^i_t \cup \{ s( u ) \mid u : \upsilon \in S^i, \upsilon \in \Data, \upsilon \text{ finite}, s \in \Sel^{\upsilon}  \}$,
and let $S^\ast_t$ be the fixed point of this sequence.
We know this is a finite set for each $t$ since the values of $t$'s type are of finite size.
Let $S^\ast$ be the union of $S^\ast_t$ for all $t \in T$,
and let $\tcEc$ be the set of subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) ) \mid t : \tau \in S \cup S^\ast, C_i \in \Ctr^\tau \}$.
In a derivation tree with root node $\Ec$, 
it can be shown by case analysis on the rules of the calculus that each non-root node $\Fc$ is such that 
$\mathcal{T}(\Fc) \subseteq \tcEc$, and moreover thus contains an equality between two terms from $\tcEc$ not occurring in its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

\begin{lemma}[Refutation Soundness]
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Unify$_2$}, or \rn{Cyclic},
then $\Ec$ is unsatisfiable in $\thD$.
\rem{Elaborate on cyclic?}
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, \rn{Unify$_1$}, \rn{Bisimilar}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
\rem{Elaborate on Bisimilar?}
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and thus $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}

@@

\begin{lemma}[Solution Soundness]
If there exists a derivation tree with root node $\Ec$ containing a saturated node $\Fc$, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
Let $\Fc$ be a saturated node in a derivation tree with root node $\Ec$.
We assume that:
\begin{enumerate}
\item Each equivalence class of $\Fc$ contains at least one variable,
\item Each equivalence class $\ec{x}$ of $\Fc$ where $x : \tau$, $\tau \in \Nondata$ contains a constant,
\item Each equivalence class $\ec{x}$ of $\Fc$ where $x : \tau$, $\tau \in \Data \cup \Codata$ contains a constructor term.
\end{enumerate}
The first statement comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability.
The second statement can be assumed for $x : \tau$, since we know that $\tau$ is infinite.
The third statement can be assumed by \rem{TODO}.

We construct a model $\M$ of $\thD$ that interprets constructors and correctly-applied selectors in the obvious way,
interprets wrongly-applied selectors based on information in $\Fc$,
interprets the variables and $x$ of $\Fc$ based on a valuation $\Val$, which is constructed as follows.
For each equivalence class $e : \tau$, where $\tau \in \Nondata$ containing constant $c$,
$\Val(x) = c$ for each $x \in e$.
For each $e : \tau$, where $\tau \in \Data$ containing the constructor term $C( \vec u )$ where the equivalence classes $\ec{\vec u}$ contains the variables $\vec y$,
$\Val(x) = C( \Val( \vec y ) )$ for each $x \in e$.
Since \rn{Cyclic} does not apply, this definition is well-defined. 
For each $e : \tau$, where $\tau \in \Codata$, 
$\Val( x ) = \interp{ \tpath{\Fc}{e} }$ for each $x \in e$.

Since \rn{Cong} does not apply, we have that:
\begin{eqnarray} \label{eqn:m-bij-dt}
\Val(x) = \Val(y) \text{ if and only if } \ec{x} = \ec{y}
\end{eqnarray}

@@

If $\tau \in \Data$, then assume the equivalence class of $t_1$ and $t_2$ contains variable $x$.
We show that $\M( t ) = \Val( x )$ for each term $t \in \ec{x}$, which implies $\M \models t_1 \teq t_2$.
In the case where $t$ is a variable, then $\M( t ) = \Val( x )$ since $t$ and $x$ are assigned identical values by construction.
In the case where $t$ is a constructor term $C( \vec u )$, say our construction built a valuation for $x$ based on the constructor term $C'( \vec u' ) \in \ec{x}$.
Since \rn{Unify$_2$} does not apply to $\Fc$, we have that $C = C'$.
Since \rn{Unify$_1$} does not apply to $\Fc$, we have that $\ec{\vec u} = \ec{\vec u'}$.
By structural induction, we have $\M( \vec u ) = \M( \vec u' )$, and thus $\M( t ) = \M( C( \vec u ) ) = \M( C'( \vec u ' ) ) = \Val( x )$.
In the case where $t$ is a selector term $s^j_k( u )$,
then since \rn{Split} does not apply, 
$\ec{u}$ must contain a term of the form $C_i( s^1_i( u ), \ldots, s^{a_i}_i( u ) )$ for some $i$.
Let $x_1, \ldots, x_n$ be the variables in equivalence classes $\ec{s^1_i( u )}, \ldots, \ec{s^{a_i}_i( u )}$.
By our construction, $\M( u ) = C_i( \Val( x_1 ), \ldots, \Val( x_{a_i} ) )$.
If $i = k$, then $\M( t ) = \M( s^j_k( u ) ) = s^j_k( \M( u ) ) = \Val( x_j )$, which is $\Val( x )$ since $x_j \in \ec{s^j_i( u )} = \ec{x}$.
If $i \neq k$, then $\thD$ does not constraint the value of $\M( t )$ and since $s^j_k( \M( u ) )$ is otherwise unconstrained in our model construction since \rn{Cong} does not apply, 
we interpret $s^j_k( \M( u ) )$ as $\Val( x )$.

@@

\rem{Thus, sound and complete.}

>>>>

A \emph{derivation tree} is a tree whose nodes are sets of equalities, where non-root nodes are obtained by 
a non-redundant application of a derivation rule to its parent node.
A derivation tree is \emph{closed} if all of its leaf nodes are $\bot$.
A node is \emph{saturated} if no non-redundant application of a rule can be applied to it.
If there exists a closed derivation tree with root node $\Ec$, then we conclude that $\Ec$ is $\thD$-unsatisfiable.
If there exists a derivation tree with root node $\Ec$ that contains a saturated node, then we conclude that $\Ec$ is $\thD$-satisfiable.

@@

\begin{lemma}[Termination] \label{lem:t}
All derivation trees are finite.
\end{lemma}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S$ be the set of terms occurring as the argument of a selector in $\tEc$.
Let $T$ be the set of terms in $\tEc$ of finite datatype type.
For each term $t \in T$,
let $S^0_t$ be the set $\{ t \}$,
for $i \geq 0$ let $S^{i+1}_t =$ $S^i_t \cup \{ s( u ) \mid u : \upsilon \in S^i, \upsilon \in \Data, \upsilon \text{ finite}, s \in \Sel^{\upsilon}  \}$,
and let $S^\ast_t$ be the fixed point of this sequence.
We know this is a finite set for each $t$ since the values of $t$'s type are of finite size.
Let $S^\ast$ be the union of $S^\ast_t$ for all $t \in T$,
and let $\tcEc$ be the set of subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) ) \mid t : \tau \in S \cup S^\ast, C_i \in \Ctr^\tau \}$.
In a derivation tree with root node $\Ec$, 
it can be shown by case analysis on the rules of the calculus that each non-root node $\Fc$ is such that 
$\mathcal{T}(\Fc) \subseteq \tcEc$, and moreover thus contains an equality between two terms from $\tcEc$ not occurring in its parent node.
Thus, the depth of a branch in a derivation tree with root node $\Ec$ is at most $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

\begin{lemma}[Refutation Soundness] \label{lem:rs}
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Unify$_2$}, or \rn{Cyclic},
then $\Ec$ is unsatisfiable in $\thD$.
\rem{Elaborate on cyclic?}
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, \rn{Unify$_1$}, \rn{Bisimilar}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
\rem{Elaborate on Bisimilar?}
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and thus $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}

@@

\begin{lemma}[Solution Soundness] \label{lem:ss}
If there exists a derivation tree with root node $\Ec$ containing a saturated node $\Fc$, then $\Ec$ is satisfiable in $\thD$.
\end{lemma}
\begin{proof}
Let $\Fc$ be a saturated node in a derivation tree with root node $\Ec$.
We assume that:
\begin{enumerate}
\item Each equivalence class of $\Fc$ contains at least one variable,
\item Each equivalence class $\ec{x}$ of $\Fc$ where $x : \tau$, $\tau \in \Nondata$ contains a constant,
\item Each equivalence class $\ec{x}$ of $\Fc$ where $x : \tau$, $\tau \in \Data \cup \Codata$ contains a constructor term.
\end{enumerate}
The first statement comes with no loss of generality since new equalities of the form $y \teq t$ for fresh variable $y$ can be added to $\Fc$ without affecting its satisfiability.
The second statement can be assumed for $x : \tau$, since $\tau$ is infinite.
Hence, we can assign constants for each such equivalence class.
Moreover, since $\Fc$ contains $\thD$-constraints only, these constants can be chosen arbitrarily.
The third statement can be assumed by \rem{TODO}.

We construct a model $\M$ of $\thD$ that interprets constructors and correctly-applied selectors in the obvious way,
interprets wrongly-applied selectors based on equality information in $\Fc$,
interprets the variables and $x$ of $\Fc$ based on a valuation $\Val$, which is constructed as follows.
For each equivalence class $e : \tau$, where $\tau \in \Nondata$ containing constant $c$,
$\Val(x) = c$ for each $x \in e$.
For each $e : \tau$, where $\tau \in \Data$ containing the constructor term $C( \vec u )$ where the equivalence classes $\ec{\vec u}$ contains the variables $\vec y$,
$\Val(x) = C( \Val( \vec y ) )$ for each $x \in e$.
Since \rn{Cyclic} does not apply, this definition is well-defined. 
For each $e : \tau$, where $\tau \in \Codata$, 
$\Val( x ) = \interp{ \tpath{\Fc}{e} }$ for each $x \in e$.
Since neither \rn{Cong} or \rn{Bisimilar} applies to $\Fc$, we have that for all variables $x$ and $y$:
\begin{eqnarray} \label{eqn:m-bij-dt}
\Val(x) = \Val(y) \text{ if and only if } \ec{x} = \ec{y}
\end{eqnarray}

@@

If $\tau \in \Data$, then assume the equivalence class of $t_1$ and $t_2$ contains variable $x$.
By induction on the size of $\Val( x )$, we show that $\M( t ) = \Val( x )$ for each term $t \in \ec{x}$, which implies $\M \models t_1 \teq t_2$.
In the case where $t$ is a variable, then $\M( t ) = \Val( x )$ since $t$ and $x$ are assigned identical values by construction.
In the case where $t$ is a constructor term $C( \vec u )$, say our construction built a valuation for $x$ based on the constructor term $C'( \vec u' ) \in \ec{x}$.
Since \rn{Unify$_2$} does not apply to $\Fc$, we have that $C = C'$.
Since \rn{Unify$_1$} does not apply to $\Fc$, we have that $\ec{\vec u} = \ec{\vec u'}$.
By the induction hypothesis, we have $\M( \vec u ) = \M( \vec u' )$, and thus $\M( t ) = C( \M( \vec u ) ) = C'( \M( \vec u ' ) ) = \Val( x )$.
In the case where $t$ is a selector term $s^j_k( u )$,
then since \rn{Split} does not apply, 
$\ec{u}$ must contain a term of the form $C_i( s^1_i( u ), \ldots, s^{a_i}_i( u ) )$ for some $i$.
Let $x_1, \ldots, x_n$ be the variables in equivalence classes $\ec{s^1_i( u )}, \ldots, \ec{s^{a_i}_i( u )}$.
By our construction, $\M( u ) = C_i( \Val( x_1 ), \ldots, \Val( x_{a_i} ) )$.
If $i = k$, then $\M( t ) = \M( s^j_k( u ) ) = s^j_k( \M( u ) ) = \Val( x_j )$, which is $\Val( x )$ since $x_j \in \ec{s^j_i( u )} = \ec{x}$.
If $i \neq k$, then $\thD$ does not constraint the value of $\M( t )$ and since $s^j_k( \M( u ) )$ is otherwise unconstrained in our model construction since \rn{Cong} does not apply, 
we may interpret $s^j_k( \M( u ) )$ as $\Val( x )$.

@@

By Lemmas~\ref{lem:t},~\ref{lem:rs}, and~\ref{lem:ss}, our calculus is sound and complete for $\thD$.
